---
title: "Getting started with BriDGE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with BriDGE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

## Why BriDGE?

Randomized controlled trials (RCTs) tell you *whether* an intervention works.
They are rarely designed to tell you *why and how* it works — which mediators
carry the effect, whether relationships are nonlinear, and whether your
theoretical causal model is actually consistent with the data.

BriDGE implements the protocol described in the companion paper
(Veltri & Banerjee, in press, *Behavior Research Methods*), combining:

1. **A researcher's DAG** — your theoretical causal model of the experiment.
2. **Causal discovery** — structure learning constrained by experimental
   design facts (the randomized treatment can affect other variables, but
   nothing can cause the treatment).
3. **GAM-based mediation analysis** — direct and indirect effects estimated
   with flexible smooth functions instead of rigid linear terms.
4. **Bootstrap inference and sensitivity checks** — so effects come with
   uncertainty and robustness information.

This vignette walks through the full workflow on synthetic data where the
ground truth is known.

```{r setup, message = FALSE}
library(BriDGE)
```

## 1. Data

`bridge_generate_data()` generates a synthetic two-arm RCT with a binary
`treatment`, two mediators, and a continuous `outcome`. The data-generating
process deliberately includes nonlinearities (a quadratic mediator effect,
a mediator–mediator interaction, and treatment-by-mediator interactions),
mirroring the simulation study in the paper.

```{r data}
data <- bridge_generate_data(n = 500, nonlinear_strength = 0.5, seed = 42)
head(data)
```

With your own data you need: a **two-level treatment factor** (the *first*
factor level is taken as control, the second as treated), numeric mediators,
a numeric outcome, and no missing values in the analysis variables:

```{r own-data, eval = FALSE}
your_data$treatment <- factor(your_data$group,
                              levels = c("control", "treatment"))
your_data <- na.omit(your_data)
```

## 2. Causal discovery: does the data agree with your DAG?

`bridge_discover()` learns a DAG from the data while enforcing what we know
by design: edges from the treatment to each mediator are whitelisted
(randomization guarantees the treatment is upstream), and any edge *into*
the treatment is blacklisted.

```{r discover}
discovery <- bridge_discover(
  data      = data,
  treatment = "treatment",
  mediators = c("mediator_1", "mediator_2"),
  outcome   = "outcome",
  method    = "mmhc"   # also: "hc", "pc"
)

discovery$comparison
```

Each row is an edge; `Researcher` marks edges in your assumed DAG (treatment
to each mediator and the outcome, each mediator to the outcome), `Discovered`
marks edges the algorithm found. Read disagreements as **hypotheses to
examine, not verdicts**: an edge the algorithm finds between mediators, for
example, suggests your parallel-mediator theory may miss a mediator-to-mediator
pathway. Discovery on discretized data is also sensitive to binning choices,
so treat low-stability edges with caution (see the paper's discussion of
arc-strength stability).

You can plot both structures:

```{r plot-dags, eval = FALSE}
plot(discovery$researcher_dag)     # your theory
plot(discovery$discovered_igraph)  # what the data suggests
```

## 3. Mediation analysis with GAMs

`bridge_mediate()` decomposes the treatment effect using counterfactual
prediction from an outcome GAM with smooth terms for each mediator:

- **`direct_effect`** — the natural direct effect: treatment's effect on the
  outcome holding mediators at their control-condition values.
- **`nie_<mediator>`** — the indirect effect through each mediator: the
  outcome change produced by shifting *that* mediator (and only that one)
  from its control to its treated value.
- **`total_effect`** — treatment's overall effect with all mediators
  responding.

Confidence intervals come from bootstrap resampling. We use a small number of
bootstraps here to keep the vignette fast — use **500–2000** in real analyses.

```{r mediate}
mediation <- bridge_mediate(
  data         = data,
  treatment    = "treatment",
  mediators    = c("mediator_1", "mediator_2"),
  outcome      = "outcome",
  n_bootstraps = 50,      # use >= 500 in practice
  nonlinear    = TRUE,    # GAM smooths for mediator-outcome links
  parallel     = FALSE,   # TRUE recommended on your machine
  seed         = 42
)

for (effect in names(mediation$summaries)) {
  s <- mediation$summaries[[effect]]
  cat(sprintf("%-16s % .3f  [% .3f, % .3f]\n",
              effect, s$mean, s$ci_lower, s$ci_upper))
}
```

An interval that excludes zero indicates a reliable effect at the 95% level.
Because the outcome model uses smooth terms, these estimates remain valid
when mediator–outcome relationships are curved — exactly the situation where
classic linear mediation (e.g., product-of-coefficients) can be biased.
Setting `nonlinear = FALSE` re-runs everything with linear models, which is a
useful comparison: if the two disagree, the nonlinearity matters.

## 4. Sensitivity analysis

`bridge_sensitivity()` adds small Gaussian noise to the mediators and outcome
and re-runs the mediation. If your conclusions change under tiny
perturbations, they were fragile to begin with.

```{r sensitivity}
sensitivity <- bridge_sensitivity(
  data            = data,
  treatment       = "treatment",
  mediators       = c("mediator_1", "mediator_2"),
  outcome         = "outcome",
  n_bootstraps    = 25,
  perturbation_sd = 0.1,
  parallel        = FALSE,
  seed            = 42
)

round(sapply(sensitivity$perturbed_results$summaries, `[[`, "mean"), 3)
```

Compare these means with the unperturbed ones above: they should be close.

## 5. Everything at once

`bridge_analyze()` runs the whole pipeline — discovery, mediation, group
comparison, sensitivity, plots, and a text summary:

```{r analyze}
results <- bridge_analyze(
  data         = data,
  treatment    = "treatment",
  mediators    = c("mediator_1", "mediator_2"),
  outcome      = "outcome",
  n_bootstraps = 50,     # use >= 500 in practice
  parallel     = FALSE,
  sensitivity  = FALSE,  # enabled by default; skipped here for speed
  seed         = 42
)

summary(results)
```

The result object keeps every component: `results$discovery`,
`results$mediation`, `results$comparison`, `results$plots`.

```{r plots}
results$plots$effect_sizes
```

```{r plots2}
results$plots$bootstrap_distributions
```

## 6. Practical guidance

- **Sample size**: mediation estimands need more power than average treatment
  effects; aim for hundreds of observations, and treat discovery output on
  small samples as exploratory.
- **Bootstraps**: 500 minimum for stable percentile intervals; more for
  publication.
- **Convergence**: on small bootstrap samples GAMs occasionally struggle.
  `handle_convergence = "simplify"` automatically falls back to simpler
  models; `"error"` makes failures loud instead.
- **Reproducibility**: pass `seed =` to make bootstrap draws reproducible,
  and report it.
- **When simpler tools suffice**: with one mediator, no interaction or
  nonlinearity suspected, classic linear mediation may be all you need — see
  the decision flowchart in the companion paper (Figure 1).

## Reproducing the paper

The full simulation scripts behind the companion paper (2-, 3-, and
5-mediator scenarios, BCa intervals, arc-strength stability, benchmarking and
power grids, and the JOBS II semi-synthetic benchmark) ship with the package:

```{r sims, eval = FALSE}
list.files(system.file("simulations", package = "BriDGE"))
```

## Citation

```{r citation, eval = FALSE}
citation("BriDGE")
```

> Veltri, G. A., & Banerjee, S. (in press). BriDGE the gap – improving
> Behavioural research by integrating DAGs and GAMs in Experiments.
> *Behavior Research Methods.*
